---
title: "Human Freedom Index and Suicides Project"
subtitle: "Part 2: Data Joining and Modeling"
author: "Doug Cady"
date: "November 13, 2021"
output:
    html_document:
        toc: true
        theme: simplex
        highlight: tango
    pdf_document:
        toc: true
    github_document:
        toc: true
---

R version 4.1.1 "Kick Things"

```{r packs, results = 'hide', message = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(psych)
# library(stringr)
# library(GGally)
# library(plotly)
```


## Load datasets into R

```{r data, results = 'hide', message = FALSE}
hfi <- read_csv("../data/clean_hfi_2018.csv")
suicides <- read_csv("../data/clean_suicides.csv")
```


## Join datasets on `year` and `country` variables
```{r initial_join}
# How many countries in each dataset?
hfi_countries <- unique(hfi$country)
print(length(hfi_countries))  # 162

suic_countries <- unique(suicides$country)
print(length(suic_countries))  # 101

init_joined <- inner_join(hfi, suicides, by = c("year", "country"))
init_join_countries <- unique(init_joined$country)
print(length(init_join_countries))  # 78

# 23 countries weren't joined together initially, why?
nonjoin_countries <- setdiff(suic_countries, init_join_countries)
print(nonjoin_countries)

# print(setdiff(hfi_countries, init_join_countries))
```


### Investigate country names that don't match
Use duckduckgo and wikipedia to sort out name mismatches (if possible)

```{r explore_non_matches}
# Antigua and Barbuda - only in suicides (ok)
# Aruba - only in suicides (ok)
# Azerbaijan - country in both, but years don't align (ok)
# Cabo Verde - in hfi named Cape Verde, change to Cape Verde in suicides data (!fix suicides)
# Cuba - only in suicides data (ok)
# Czech Republic - hfi abbreviated to Czech Rep. (!fix hfi)
# Dominica - only in suicides, different from Dominican Republic (in hfi) (ok)
# Grenada - only in suicides (ok)
# Kiribati - only in suicides (ok)
# Kyrgyzstan - hfi lists official name "Kyrgyz Republic" (!fix hfi)
# Macau - only in suicides (ok)
# Maldives - only in suicides (ok)
# Puerto Rico - only in suicides (ok)
# Republic of Korea - hfi lists as "Korea, South" (!fix hfi)
# Russian Federation - hfi lists as "Russia" (!fix suicides)
# Saint Kitts and Nevis - only in suicides (ok)
# Saint Lucia - only in suicides (ok)
# Saint Vincent and Grenadines - only in suicides (ok)
# San Marino - only in suicides (ok)
# Slovakia - hfi lists as "Slovak Rep." (!fix hfi)
# Sri Lanka - country in both, but years don't align (ok)
# Turkmenistan - only in suicides (ok)
# Uzbekistan - only in suicides (ok)

## Check existence of country and years available in each dataset:
# HFI
print(unique(filter(hfi, str_detect(country, "India"))$country))
print(unique(filter(hfi, str_detect(country, "India"))$year))
# Suicides
print(unique(filter(suicides, str_detect(country, "India"))$country))
print(unique(filter(suicides, str_detect(country, "India"))$year))
```


### Clean up country names that didn't match initially
```{r clean_country_names}
hfi_clean_cntry <- hfi %>%
    mutate(country = str_replace(country, "Czech Rep.", "Czech Republic"),
           country = str_replace(country, "Kyrgyz Republic", "Kyrgyzstan"),
           country = str_replace(country, "Korea, South", "Republic of Korea"),
           country = str_replace(country, "Slovak Rep.", "Slovakia"),)


suic_clean_cntry <- suicides %>%
    mutate(country = str_replace(country, "Cabo Verde", "Cape Verde"),
           country = str_replace(country, "Russian Federation", "Russia"))

joined <- inner_join(hfi_clean_cntry, suic_clean_cntry, by = c("year", "country"))

print(length(unique(joined$country))) # 78 init, 84 final
```

``` {r year_explore_counts}
# 2016 doesn't seem to have many records, maybe it should be excluded from this analysis?
# Yes, it has far fewer (< 20%) records than other years have
joined %>%
    group_by(year) %>%
    count()

joined_no2016 <- filter(joined, year != 2016)
write_csv(joined_no2016, "../output/clean_hfi_suicides_joined_no2016.csv")
```


## Explore correlations between freedom, suicides, and GDP
Overall, personal, and economic freedom scores, number of suicides per 100k, and GDP per capita

```{r explore_correlations, fig.height = 7, fig.width = 10, cache = TRUE}
# Suicides are currently broken down by age and sex, we need to aggregate to total counts
corr_data <- joined_no2016 %>%
    group_by(year, country) %>%
    mutate(total_suicides = sum(suicides_no),
           total_pop = sum(population),
           total_suicides_p100k = total_suicides / total_pop * 100000) %>%
    ungroup() %>%
    select(hf_score, pf_score, ef_score, total_suicides_p100k, gdp_per_capita)

summary(corr_data)

# Correlation pairs plot with histograms, density, and scatter plots
pairs.panels(corr_data,
             smooth = TRUE,
             scale = FALSE,
             density = TRUE,
             ellipses = FALSE,
             method = "pearson",
             pch = 21,
             lm = FALSE,
             cor = TRUE,
             jiggle = FALSE,
             hist.col = 4,
             stars = TRUE,
             ci = TRUE)
```

#### Overall human freedom score seems to be most highly correlated with GDP per capita, though this is not too surprising.  Suicide rates do seem to have a weak correlation to personal and human freedom, but not economic freedom or gdp per capita.  This makes me wonder if they would be more corelated to a country's happiness rather than it's freedom.



## Does a country's population size impact freedom?
#### No, there doesn't seem to exist any relationship between population size and human freedom
```{r pop_size_freedom, fig.height = 7, fig.width = 10, warning = FALSE, message = FALSE}
ggplot(joined_no2016, aes(x = population, y = hf_score)) +
    geom_point() +
    geom_smooth() +
    facet_wrap( ~ year, nrow = 2)
```


## Do female freedom scores relate to female suicide rates?
```{r fem_freedom_suicide, fig.height = 7, fig.width = 10, warning = FALSE, message = FALSE}
fem_freedom_suic <- joined_no2016 %>%
    filter(sex == 'female') %>%
    group_by(year, country) %>%
    mutate(total_suicides = sum(suicides_no),
           total_pop = sum(population),
           total_suicides_p100k = total_suicides / total_pop * 100000) %>%
    ungroup() %>%
    select(year, country, pf_ss_women_fgm, pf_ss_women_inheritance, pf_movement_women,
           pf_identity_sex_female, pf_identity_divorce, total_suicides_p100k)

fem_freedom_suic$avg_fem_freedom <- rowMeans(select(fem_freedom_suic,
                                                    pf_ss_women_fgm, pf_ss_women_inheritance,
                                                    pf_movement_women, pf_identity_sex_female,
                                                    pf_identity_divorce), na.rm = TRUE)

ggplot(fem_freedom_suic, aes(x = avg_fem_freedom, y = total_suicides_p100k)) +
    geom_jitter(alpha = 0.1) +
    geom_smooth(se = FALSE)
```

#### The relationship between female freedom in a country and female suicide rates is non-existent to weak in strength. Many countries have top freedom scores and have a wide range of suicide rates per one hundred thousand persons.


## What are important factors that contribute to human freedom?
```{r freedom_factors}



```
